One of the most significant changes introduced in version 4 was the use of
Python.  The input file is actually a Python script, which interacts with a Psi
Python module to perform computations.  In order for this to happen, the C++
binding must be known to Python; this is all done in the 
{\tt \$PSI4/src/bin/psi4/python.cc} file.  For example, we have an SCF module, with
the C++ signature
{\tt PsiReturnType cscf::cscf(Options \&options);}
To allow Python to use this, we first define a little wrapper function
\begin{verbatim}
double py_psi_scf()
{
    if (scf::scf(Process::environment.options) == Success)
        return Process::environment.globals["CURRENT ENERGY"];
    else
        return 0.0;
}
\end{verbatim}
This does a couple of things to automate things a) it passes the default
options object into SCF automatically, so that the user doesn't have to, and b)
checks the return value, and will return the energy, which is posted to the
globals map, back to Python.  Note that this is C++ code, within \PSIfour\ so
it is aware of all global objects, such as PSIO, Chkpt and Options.  Now we
have this simple function call, we can tell Python about it:
\begin{verbatim}
def("scf",  py_psi_scf);
\end{verbatim}
This binds the keyword ``scf'' to the newly created wrapper function, allowing
the user to type ``scf()'' in their Python input file to fire up the SCF
module.  Similarly, the user might want to be able to call {\tt Molecule}'s
member functions directly from Python.  This can also be done easily:

\begin{verbatim}
class_<Molecule, shared_ptr<Molecule> >("Molecule").
        def("print_to_output", &Molecule::print).
        def("nuclear_repulsion_energy", &Molecule::nuclear_repulsion_energy);
\end{verbatim}

This first defines the keyword {\tt Molecule} to refer to the C++ {\tt
Molecule}; the {\tt shared\_ptr<Molecule>} keyword tells Python to store it as
a shared pointer, which ensures that the object will not be deleted until both
C++ and Python have no more references to it.  The member functions to be bound
are then specified by a chained sequence of {\tt def} calls (note the periods),
terminated by a semicolon.  Then, if the user had defined a molecule called
``water'', they could print its geometry simply with the command {\tt
water.print\_to\_output()}.

Direct interaction with the Psi module from Python requires function calls that
look like {\tt PsiMod.call\_some\_function()}.  This is not very friendly to
your average user, so a preprocessor checks for known Psi syntax and turns it
into valid Python, before handing it off for excecution.  This preprocessor is
purely Python, and lives in {\tt \$PSI4/lib/python/input.py}.  For example, the
following text \begin{verbatim}
set scf {
    SCF_TYPE DIRECT
    BASIS cc-pVDZ
    RI_BASIS_SCF cc-pVDZ-HF
    guess core
}
\end{verbatim}
is converted to the following text
\begin{verbatim}
PsiMod.set_default_options_for_module("SCF")
PsiMod.set_option("SCF_TYPE", "DIRECT")
PsiMod.set_option("BASIS", "cc-pVDZ")
PsiMod.set_option("RI_BASIS_SCF", "cc-pVDZ-HF")
PsiMod.set_option("GUESS", "core")
\end{verbatim}
which can be handled by Python.

There are a number of other utilities, which are entirely Python, located in
{\tt \$PSI4/lib/python}.  These provide convenient functions to the user, such
as {\tt table}.

